---
title: "calluna_xylem_cleaning"
format: html
editor: visual
---

# Calluna xylem cleaning

## Setup

```{r}
#| label: libraries
#| message: false
#| warning: false

library(tidyverse)
library(ggplot2)
library(patchwork)
library(lme4)
library(plotrix)
```

```{r}
#| label: import

raw_data <- read.csv("raw_data/DURIN_raw_LygraFire_field_trait_2025.csv")
raw_data$age_class <-factor(raw_data$age_class, levels = c("PIO", "BUI", "MAT"))

```

## Cleaning

### Evenness

```{r}
#| label: check evenness
#| message: false
#| warning: false

#this chunk creates a dataframe that count the amount of xylem per treatment before and after filtering by circularity

xylem_count_evenness <- raw_data |>
  count(treatment)

xylem_count_evenness_filtered <- raw_data |>
  mutate(circ = (4*3.14)*(Area/(Perim.*Perim.))) |>
  filter(circ >= 0.70) |>
  count(treatment)

xylem_count_combined <- xylem_count_evenness |>
  mutate(count_filter = xylem_count_evenness_filtered$n) |>
  mutate(percentage_remaining = count_filter/n)

names(xylem_count_combined) <- c("Treatment", "Number of xylems", "Number of xylems after filtering", "Percentage xylem remaining")
  

xylem_count_combined
```

```{r}
#| label: circularity in relation to damage

#this chunk provides a comparison of datapoints before and after filtering by circularity >= 0.70.

data_circ <- raw_data |>
  mutate(circ = (4*3.14)*(Area/(Perim.*Perim.))) |>
  group_by(treatment) |>
  mutate(means_treatment = mean(Area))

data_circ_filter <- data_circ |>
  filter(circ >= 0.70)

#creating new filtered datasets
xylem_damage_5 <- data_circ |>
  filter(subset_damage == 5)

xylem_damage_4 <- data_circ |>
  filter(subset_damage == 4)

xylem_damage_5_filtered <- data_circ_filter |>
  filter(subset_damage == 5)

xylem_damage_4_filtered <- data_circ_filter |>
  filter(subset_damage == 4)

#create a custom palette to use
palette_damage <- c("orange", "red")


plot_damage <- ggplot() + 
  geom_jitter(data = data_circ, aes(x = treatment, y = Area), color = "grey") + geom_jitter(data = xylem_damage_4, aes(x = treatment, y = Area, color = as.character(as.numeric(subset_damage)))) + geom_jitter(data = xylem_damage_5, aes(x = treatment, y = Area, color = as.character(as.numeric(subset_damage)))) + geom_point(data = data_circ, aes(x = treatment, y = means_treatment), color = "blue") + labs(title = "Unfiltered") + ylab(label = "Xylem vessel area in µm") + xlab(label = "NiN x Age class") + theme(legend.position = "none") + scale_color_manual(values = palette_damage) + scale_y_continuous(breaks = seq(100,2000, by = 100))

plot_damage_filtered <- ggplot() + 
  geom_jitter(data = data_circ, aes(x = treatment, y = Area), color = "grey") + geom_jitter(data = xylem_damage_4_filtered, aes(x = treatment, y = Area, color = as.character(as.numeric(subset_damage)))) + geom_jitter(data = xylem_damage_5_filtered, aes(x = treatment, y = Area, color = as.character(as.numeric(subset_damage)))) + geom_point(data = data_circ, aes(x = treatment, y = means_treatment), color = "blue") + labs(title = "Circularity >= 0.7") + ylab(label = "Xylem vessel area in µm") + xlab(label = "NiN x Age class") + scale_color_manual(values = palette_damage) + guides(color = guide_legend(title = "Subset damage index")) + scale_y_continuous(breaks = seq(100,2000, by = 100))

plot_damage + plot_damage_filtered
```

NOTE: blue dots indicate treatment level averages (e.g. BUI-TH)

The patterns of means for every treatment remain largely the same after filtering by circularity \>= 0.70, implying an uniform distribution of damaged and misshapen xylem. In this case, almost all of the xylem from subsets with damage index 5 is removed, while a large portion of xylem from subsets with damage index 4 is removed too. Circularity seems to be a good rule for excluding structures that are likely not xylem or are xylem damaged during lab work. Filtering by circularity \>= 0.70 removes 4000 datapoints, which is approximately 25% of the total data set.

```{r}
#| label: assessing skewness after filtering

skew_raw <- raw_data |>
  mutate(set = "raw")

skew_filtered <- data_circ_filter |>
  mutate(set = "filter") |>
  select(-c(24,25))

skew_both <- rbind(skew_raw, skew_filtered)

skew_summary <- skew_both |>
  group_by(set, age_class, nin_type) |>
  summarise(
    skew = moments::skewness(Area),
    mean = mean(Area),
    sd = std.error(Area),
    n = n()
  ) |> 
  ungroup() |>
  mutate(treatment = paste(age_class, nin_type))


skew_plot <- ggplot(data = skew_summary) + geom_point(aes(x = treatment, y = skew, color = set))

mean_plot <- ggplot(data = skew_summary) + geom_point(aes(x = treatment, y = mean, color = set)) + geom_errorbar(aes(x = treatment, ymin = mean-sd, ymax = mean+sd, color = set))

skew_plot + mean_plot
```

A reason for the increased size is the presence of double parenchyma cells, small tears, and stretched large parenchyma / small xylems, which is gets filtered out by the circularity filter. Many of these are below the average xylem size and thus seemingly increases the average xylem area. The filter also removes large xylem vessels, but they are not as common.\
\
MAT-TH actually decreases in average size after filtering, but other than that there is a shared tendency to increase average size after filtering.

```{r}
#| label: chi square test

data_chi <- raw_data |>
  mutate(circ = (4*3.14)*(Area/(Perim.*Perim.))) |>
  mutate(kept = case_when(
    circ <= 0.70 ~ 0,
    circ >= 0.70 ~ 1
  )) |> 
  group_by(treatment) |>
  count(kept)

data_chi <- xtabs(n ~ treatment + kept, data = data_chi)

chisq.test(data_chi)

```

The chi squared test shows that the filtering rule affects each treatment differently, so it is important to quantify if this bias is acceptable or favors a particular treatment. BUI-BH had the highest reduction of data points and was one of the treatments with largest change in mean area.

```{r}
problem_treatment <- raw_data |>
  mutate(circ = (4*3.14)*(Area/(Perim.*Perim.))) |>
  mutate(kept = case_when(
    circ <= 0.70 ~ 0,
    circ >= 0.70 ~ 1
  )) |> 
  subset( kept == 0 & treatment == "BUI-BH") |>
  select(c(2,3,4,24))


```

```{r}
model_mixed_raw <- lmer(Area ~ nin_type * age_class + (1 | plant), data = raw_data)
#model_mixed <- lmer(Area ~ nin_type * age_class + (1 | plant), data = filtered_data)

res_raw <- residuals(model_mixed_raw)
#res <- residuals(model_mixed)

qqnorm(res_raw)
qqline(res_raw)
#qqnorm(res)
#qqline(res)

plot(model_mixed_raw)
```

The plot above shows the difference between the residuals of the unfiltered

```{r}

model_mixed_raw_log <- lmer(log(Area) ~ nin_type * age_class + (1 | plant), data = raw_data)
model_mixed_log <- lmer(log(Area) ~ nin_type * age_class + (1 | plant), data = filtered_data)

res_raw_log <- residuals(model_mixed_raw_log)
res_log <- residuals(model_mixed_log)


qqnorm(res_raw_log)
qqline(res_raw_log)
qqnorm(res_log)
qqline(res_log)
```

```{r}
plot(res_raw)
plot(res_raw_log)
```

Due to the data being skewed, transforming the data logarithmically improves the skewness and allows the residuals to spread more evenly along the fitted line.

```         
```

```{r}
#| label: weighted circularity model

a <- 2 #this the amplitude of the weights
b <- 0.70 #this is the threshold for circularity in which it is more likely to be a xylem. everything below can easily be a tear. 

weighted_circ_data <- raw_data |>
  mutate(circ = (4*3.14)*(Area/(Perim.*Perim.))) |> #calculate circularity using imageJ's formula
  mutate(weighted_circ = circ * (a * max(0, circ - b)) * (1/(1 + subset_damage)))  |> #circularity is multiplied so that higher values of circularity has higher influence on the model versus lower values. the last part is the subset_damage index, basically higher numbers mean higher chance that the xylem are damaged or just tears
  group_by(plant) |>
  mutate(plant_mean = mean(Area)) 


mod <- lmer(Area ~ age_class * nin_type + (1 | source_file), data = weighted_circ_data, weights = weighted_circ)
mod2 <- lmer(Area ~ age_class * nin_type + (1 | plant), data = weighted_circ_data, weights = weighted_circ)
mod3 <- lmer(Area ~ age_class * nin_type + (1 | plant/source_file), data = weighted_circ_data, weights = weighted_circ)

plot(mod)
plot(mod2)
plot(mod3)
summary(mod)
summary(mod2)
summary(mod3)


res_mod <- residuals(mod)
plot(res_mod)
qqnorm(res_mod)
qqline(res_mod)

res_mod2 <- residuals(mod2)
plot(res_mod2)
qqnorm(res_mod2)
qqline(res_mod2)

res_mod3 <- residuals(mod3)
plot(res_mod3)
qqnorm(res_mod3)
qqline(res_mod3)
```

### Export

```{r}
write.csv(weighted_circ_data, "clean_data/DURIN_clean_LygraFire_field_trait_2025.csv")
```
