---
title: "calluna_xylem_cleaning"
format: html
editor: visual
---

# Calluna xylem cleaning

## Setup

```{r}
#| label: libraries
#| message: false
#| warning: false

library(tidyverse)
library(ggplot2)
library(patchwork)
library(lme4)
library(plotrix)
library(readxl)
```

```{r}
#| label: import

raw_data <- read.csv("raw_data/DURIN_raw_LygraFire_field_trait_2025.csv")
raw_data$age_class <-factor(raw_data$age_class, levels = c("PIO", "BUI", "MAT"))

```

```{r}
#| label: duplicate subset import

duplicate_xylem <- read_excel("raw_data/Duplicate_xylem_subsets.xlsx")

```

## Cleaning

### Flagging and removing duplicates

```{r}
#| label: flagging duplicate subsets

#creating the temporary cleaning dataframe
cleaning_data <- raw_data |>
  mutate(circ = (4*3.14)*(Area/(Perim.*Perim.)))  

#this creates a new column that matches duplicated subsets from the "duplicate xylem" dataframe and matches them with the proper xylem observations.
cleaning_data$duplicate_subset <- duplicate_xylem$Overlap_1[match(cleaning_data$subset_id, duplicate_xylem$Overlap_1)]

#the matching from the code above puts the name of the subset in the column, so this renames them to "1" and unaffected subsets to "0"
cleaning_data <- cleaning_data |>
  mutate(duplicate_subset = case_when(
    duplicate_subset != "NA" ~ 1,
    .default = 0
  )) |>
  filter(duplicate_subset == 0) #this line removes the duplicates


```

### Flagging old pioneers

```{r}
#| label: flagging old pioneers

#based on the cross sections of PIO-BH, they all seem too old for their burn year (>3 years), flagging them.
cleaning_data <- cleaning_data |>
  mutate(suspicious_age = case_when(
    treatment == "PIO-BH" ~ 1,
    .default = 0
  ))

```

### Distribution of damage

```{r}
#| label: distribution of damage 

damage_distribution4 <- raw_data |>
  filter(subset_damage == 4) |>
  count(treatment)

damage_distribution5 <- raw_data |>
  filter(subset_damage == 5) |>
  count(treatment)


```

### Evenness

```{r}
even_subsets <- cleaning_data |>
  group_by(
    
  ) |>
  count(subset_id)
  
```

```{r}
#| label: check evenness
#| message: false
#| warning: false

#this chunk creates a dataframe that count the amount of xylem per treatment before and after filtering by circularity

xylem_count_evenness <- raw_data |>
  count(treatment)

xylem_count_evenness_filtered <- raw_data |>
  mutate(circ = (4*3.14)*(Area/(Perim.*Perim.))) |>
  filter(circ >= 0.65) |>
  count(treatment)

xylem_count_combined <- xylem_count_evenness |>
  mutate(count_filter = xylem_count_evenness_filtered$n) |>
  mutate(percentage_remaining = count_filter/n)

names(xylem_count_combined) <- c("Treatment", "Number of xylems", "Number of xylems after filtering", "Percentage xylem remaining")
  

xylem_count_combined
```

### Damage

```{r}
damaged_subsets <- c("PIO-TH6-N")
```

```{r}
#| label: circularity in relation to damage

#this chunk provides a comparison of datapoints before and after filtering by circularity >= 0.70.

data_circ <- raw_data |>
  mutate(circ = (4*3.14)*(Area/(Perim.*Perim.))) |>
  group_by(treatment) |>
  mutate(means_treatment = mean(Area))

data_circ_filter <- data_circ |>
  filter(circ >= 0.70)

#creating new filtered datasets
xylem_damage_5 <- data_circ |>
  filter(subset_damage == 5)

xylem_damage_4 <- data_circ |>
  filter(subset_damage == 4)

xylem_damage_5_filtered <- data_circ_filter |>
  filter(subset_damage == 5)

xylem_damage_4_filtered <- data_circ_filter |>
  filter(subset_damage == 4)

#create a custom palette to use
palette_damage <- c("orange", "red")


plot_damage <- ggplot() + 
  geom_jitter(data = data_circ, aes(x = treatment, y = Area), color = "grey") + geom_jitter(data = xylem_damage_4, aes(x = treatment, y = Area, color = as.character(as.numeric(subset_damage)))) + geom_jitter(data = xylem_damage_5, aes(x = treatment, y = Area, color = as.character(as.numeric(subset_damage)))) + geom_point(data = data_circ, aes(x = treatment, y = means_treatment), color = "blue") + labs(title = "Unfiltered") + ylab(label = "Xylem vessel area in µm") + xlab(label = "NiN x Age class") + theme(legend.position = "none") + scale_color_manual(values = palette_damage) + scale_y_continuous(breaks = seq(100,2000, by = 100))

plot_damage_filtered <- ggplot() + 
  geom_jitter(data = data_circ, aes(x = treatment, y = Area), color = "grey") + geom_jitter(data = xylem_damage_4_filtered, aes(x = treatment, y = Area, color = as.character(as.numeric(subset_damage)))) + geom_jitter(data = xylem_damage_5_filtered, aes(x = treatment, y = Area, color = as.character(as.numeric(subset_damage)))) + geom_point(data = data_circ, aes(x = treatment, y = means_treatment), color = "blue") + labs(title = "Circularity >= 0.7") + ylab(label = "Xylem vessel area in µm") + xlab(label = "NiN x Age class") + scale_color_manual(values = palette_damage) + guides(color = guide_legend(title = "Subset damage index")) + scale_y_continuous(breaks = seq(100,2000, by = 100))

plot_damage + plot_damage_filtered
```

NOTE: blue dots indicate treatment level averages (e.g. BUI-TH)

The patterns of means for every treatment remain largely the same after filtering by circularity \>= 0.70, implying an uniform distribution of damaged and misshapen xylem. In this case, almost all of the xylem from subsets with damage index 5 is removed, while a large portion of xylem from subsets with damage index 4 is removed too. Circularity seems to be a good rule for excluding structures that are likely not xylem or are xylem damaged during lab work. Filtering by circularity \>= 0.70 removes 4000 datapoints, which is approximately 25% of the total data set.\
\

```{r}
#| label: damage vs circularity

damage_data <- raw_data |>
  mutate(circ = (4*3.14)*(Area/(Perim.*Perim.))) |>
  filter(subset_damage < 3)

circularity_plot <- ggplot(data = data_circ_filter) + geom_point(aes(x = circ, y= Area), alpha = 0.4)
damage_plot <- ggplot(data = damage_data) + geom_point(aes(x = circ, y= Area, color = subset_damage), alpha = 0.4)

circularity_plot + damage_plot

```

The plot above tries to illustrate if removing damage index 4 and 5 (right plot) would remove the same points as circularity \> 0.70. Removing damage 4 and 5 still keeps a lot of points between circ = 0.50 and circ = 0.70.

```{r}
#| label: potential problematic xylem

potential_problematic_xylem <- raw_data |>
  mutate(circ = (4*3.14)*(Area/(Perim.*Perim.))) |>
  group_by(source_file) |>
  mutate(subset_means = mean(Area)) |>
  mutate(problematic = case_when(
    circ <= 0.65 & Area >= (subset_means + (subset_means * 0.2)) ~ 1, #this basically checks for uncircular xylem that also are larger than the means, as stacked xylem should have a tendency to be larger than average.
    .default = 0
  )) |>
  filter(problematic == 0)

before_plot <- ggplot(data = data_circ) + geom_point(aes(x = circ, y = Area), alpha = 0.4) + facet_grid(cols = vars(age_class), rows = vars(nin_type))
after_plot <- ggplot(data = potential_problematic_xylem) + geom_point(aes(x = circ, y = Area), alpha = 0.4) + facet_grid(cols = vars(age_class), rows = vars(nin_type))

before_plot + after_plot

```

This filtering strategy is quite effective at removing stacked xylem, i have cross checked some of the pictures and it flagged just about every stacked xylem. It does not flag really jagged xylem that are close to the mean of the subset, but that could be filtered by circularity \<= 0.6 or 0.55 just to be sure to remove tears. This flagging does create discontinuity at where the threshold for circularity is set.

### Filtering by circularity \> 0.70

```{r}
#| label: assessing skewness after filtering

data_skew55 <- data_circ |>
  filter(circ > 0.55)

data_skew60 <- data_circ |>
  filter(circ > 0.60)

data_skew65 <- data_circ |>
  filter(circ > 0.65)

data_skew70 <- data_circ |>
  filter(circ > 0.70)

skew_raw <- raw_data |>
  mutate(set = "raw")

skew_filtered55 <- data_skew55 |>
  mutate(set = "filter") |>
  select(-c(24,25))

skew_filtered60 <- data_skew60 |>
  mutate(set = "filter") |>
  select(-c(24,25))

skew_filtered65 <- data_skew65 |>
  mutate(set = "filter") |>
  select(-c(24,25))

skew_filtered70 <- data_skew70 |>
  mutate(set = "filter") |>
  select(-c(24,25))

skew_both55 <- bind_rows(skew_raw, skew_filtered55)

skew_both60 <- bind_rows(skew_raw, skew_filtered60)

skew_both65 <- bind_rows(skew_raw, skew_filtered65)

skew_both70 <- bind_rows(skew_raw, skew_filtered70)

skew_summary55 <- skew_both55 |>
  group_by(set, age_class, nin_type) |>
  summarise(
    skew = moments::skewness(Area),
    mean = mean(Area),
    sd = std.error(Area),
    n = n()
  ) |> 
  ungroup() |>
  mutate(treatment = paste(age_class, nin_type))

skew_summary60 <- skew_both60 |>
  group_by(set, age_class, nin_type) |>
  summarise(
    skew = moments::skewness(Area),
    mean = mean(Area),
    sd = std.error(Area),
    n = n()
  ) |> 
  ungroup() |>
  mutate(treatment = paste(age_class, nin_type))

skew_summary65 <- skew_both65 |>
  group_by(set, age_class, nin_type) |>
  summarise(
    skew = moments::skewness(Area),
    mean = mean(Area),
    sd = std.error(Area),
    n = n()
  ) |> 
  ungroup() |>
  mutate(treatment = paste(age_class, nin_type))

skew_summary70 <- skew_both70 |>
  group_by(set, age_class, nin_type) |>
  summarise(
    skew = moments::skewness(Area),
    mean = mean(Area),
    sd = std.error(Area),
    n = n()
  ) |> 
  ungroup() |>
  mutate(treatment = paste(age_class, nin_type))

skew_plot <- ggplot(data = skew_summary) + geom_point(aes(x = treatment, y = skew, color = set))

mean_plot55 <- ggplot(data = skew_summary55) + geom_point(aes(x = treatment, y = mean, color = set)) + geom_errorbar(aes(x = treatment, ymin = mean-sd, ymax = mean+sd, color = set)) + ggtitle("Circularity > 0.55")

mean_plot60 <- ggplot(data = skew_summary60) + geom_point(aes(x = treatment, y = mean, color = set)) + geom_errorbar(aes(x = treatment, ymin = mean-sd, ymax = mean+sd, color = set)) + ggtitle("Circularity > 0.60")

mean_plot65 <- ggplot(data = skew_summary65) + geom_point(aes(x = treatment, y = mean, color = set)) + geom_errorbar(aes(x = treatment, ymin = mean-sd, ymax = mean+sd, color = set)) + ggtitle("Circularity > 0.65")

mean_plot70 <- ggplot(data = skew_summary70) + geom_point(aes(x = treatment, y = mean, color = set)) + geom_errorbar(aes(x = treatment, ymin = mean-sd, ymax = mean+sd, color = set)) + ggtitle("Circularity > 0.70")

mean_plot55 + mean_plot60 + mean_plot65 + mean_plot70
```

A reason for the increased size is the presence of double parenchyma cells, small tears, and stretched large parenchyma / small xylems, which is gets filtered out by the circularity filter. Many of these are below the average xylem size and thus seemingly increases the average xylem area. The filter also removes large xylem vessels, but they are not as common.\
\
MAT-TH actually decreases in average size after filtering, but other than that there is a shared tendency to increase average size after filtering.

```{r}
#| label: chi square test

data_chi <- raw_data |>
  mutate(circ = (4*3.14)*(Area/(Perim.*Perim.))) |>
  mutate(kept = case_when(
    circ <= 0.70 ~ 0,
    circ >= 0.70 ~ 1
  )) |> 
  group_by(treatment) |>
  count(kept)

data_chi <- xtabs(n ~ treatment + kept, data = data_chi)

chisq.test(data_chi)

```

The chi squared test shows that the filtering rule affects each treatment differently, so it is important to quantify if this bias is acceptable or favors a particular treatment. BUI-BH had the highest reduction of data points and was one of the treatments with largest change in mean area.

```{r}
problem_treatment <- raw_data |>
  mutate(circ = (4*3.14)*(Area/(Perim.*Perim.))) |>
  mutate(kept = case_when(
    circ <= 0.70 ~ 0,
    circ >= 0.70 ~ 1
  )) |> 
  subset( kept == 0 & treatment == "BUI-BH") |>
  select(c(2,3,4,24))


```

```{r}
#| label: removing damage 4 and 5

#remove datapoints that have damage level 5
cleaning_data <- cleaning_data |>
  filter(subset_damage < 4)

```

```{r}
#| label: removing circularity < 0.60

cleaning_data <- cleaning_data |>
  filter(circ > 0.60)

```

### Export

```{r}
write.csv(cleaning_data, "clean_data/DURIN_clean_LygraFire_field_trait_2025.csv")
```
