---
title: "calluna_xylem_exploration"
format: html
editor: visual
---

# Calluna xylem exploration

### Loading libraries

```{r}
#| label: libraries
#| message: false
#| warning: false

library(tidyverse)
library(readr)
library(readxl)
library(ggplot2)
library(janitor)
library(plotrix)
library(patchwork)
library(ggdist)
library(lme4)
```

### Import

```{r}
#| label: import
#| message: false
#| warning: false
#| echo: false

#the main dataset is imported in this chunk. many csv files from one directory are imported and combined into one large dataset. the dataset consists of imagej outputs of particle analyses, set with certain parameters.

#set directory. change if you use another path
xylem_area_dir <- "/home/eirik/Documents/skole/r/calluna-hydraulics/raw_data/xylem_area/"

#list each file in the xylem area directory and prepare it for import
xylem_area_files <- list.files(xylem_area_dir, pattern = "\\.csv$", full.names = TRUE)

#create a raw dataset by reading the csv files in the directory and identify them by the filename
xylem_area_raw <- xylem_area_files |>
  setNames(basename(xylem_area_files)) |>
  purrr::map_df(~ readr::read_csv(.x, show_col_types = FALSE), .id = "source_file") |>
  rename_at("...1", ~"xylem_id")
```

```{r}
#| label: subset import
#| message: false
#| warning: false
#| echo: false

#the main dataset is imported in this chunk. many csv files from one directory are imported and combined into one large dataset. the dataset consists of imagej outputs of particle analyses, set with certain parameters.

#set directory. change if you use another path
subset_area_dir <- "/home/eirik/Documents/skole/r/calluna-hydraulics/raw_data/subset_area/"

#list each file in the xylem area directory and prepare it for import
subset_area_files <- list.files(subset_area_dir, pattern = "\\.csv$", full.names = TRUE)

#create a raw dataset by reading the csv files in the directory and identify them by the filename
subset_area_raw <- subset_area_files |>
  setNames(basename(subset_area_files)) |>
  purrr::map_df(~ readr::read_csv(.x, show_col_types = FALSE), .id = "source_file") |>
  rename_at("...1", ~"subset_id")
```

```{r}
#| label: damage_import
#| message: false

#this imports the dataset that contains information about the extent of damage from microtome sectioning and growth abnormalities in the calluna that has been sampled for hydraulic analysis. damage for the whole stem, as well as damage in the "cardinal directions" and subsets.

#import damage data
xylem_damage <- read_csv("raw_data/Xylem_damage_report.csv") |>
  mutate(plant = substr(image_id,1,7))
```

\

### Combining datasets and export of raw data

```{r}
#| label: combination_prep
#creating some new column in the xylem area dataset to allow the use of match arguments so that each dataset can be combined

xylem_area_combined <- xylem_area_raw |>
  #create plant name
  mutate(plant = substr(source_file,1,7)) |>
  #create age class
  mutate(age_class = substr(source_file,1,3)) |>
  #create nin type
  mutate(nin_type = substr(source_file,5,6)) |>
  #create treatment combo
  mutate(treatment = substr(source_file, 1, 6)) |>
  #create plant cardinal direction
  mutate(plant_direction = substr(source_file,1,9)) |>
  #create own column with the subset number (to match with the damage dataset)
  mutate(subset = substr(source_file,18,18)) |>
  mutate(subset_id = paste(plant_direction,"-",subset,sep = ""))

```

```{r}
#| label: subset area combination

subset_combined <- subset_area_raw |>
  mutate(identifier = substr(source_file, 1, 9)) |>
  mutate(match_id = paste(identifier,"-",subset_id, sep = ""))

xylem_area_combined$subset_area <- subset_combined$Area[match(xylem_area_combined$subset_id, subset_combined$match_id)]

```

```{r}
#| label: combine_damage_dataset

#adding xylem damage to the combined dataset

xylem_area_combined <- xylem_area_combined |>
  mutate(subset_damage = 1) #1 is just a placeholder, will get replaced in the next chunk

#add in whole plant damage
xylem_area_combined$whole_plant_damage <- xylem_damage$damage_whole[match(xylem_area_combined$plant, xylem_damage$plant)]

#add in cardinal direction damage
xylem_area_combined$cardinal_direction_damage <- xylem_damage$damage_pic[match(xylem_area_combined$plant_direction, xylem_damage$image_id)]

#add in subset damage
xylem_area_combined$subset1 <- xylem_damage$damage_sub1[match(xylem_area_combined$plant_direction, xylem_damage$image_id)]

xylem_area_combined$subset2 <- xylem_damage$damage_sub2[match(xylem_area_combined$plant_direction, xylem_damage$image_id)]

xylem_area_combined$subset3 <- xylem_damage$damage_sub3[match(xylem_area_combined$plant_direction, xylem_damage$image_id)]

#usig the 3 new columns from above, add the respective damage from the correct subset into this column. 
xylem_area_combined <- xylem_area_combined |>
  mutate(subset_damage = case_when(
    subset == 1 ~ subset1,
    subset == 2 ~ subset2,
    subset == 3 ~ subset3
  ))


#add in warps (mostly growth abnormalities or possibly stretching from sectioning)
xylem_area_combined$warped <- xylem_damage$warped[match(xylem_area_combined$plant_direction, xylem_damage$image_id)]

#add in folds (when the section has folded over it self somewhat)
xylem_area_combined$folds <- xylem_damage$folds[match(xylem_area_combined$plant_direction, xylem_damage$image_id)]

#add in late season missing (where the outermost part of the ring has been teared off)
xylem_area_combined$late_season_missing <- xylem_damage$late_season_missing[match(xylem_area_combined$plant_direction, xylem_damage$image_id)]

#remove the 3 columns with subset damage as they are now redundant
xylem_area_combined <- xylem_area_combined |>
  select(-c(21,22,23))
```

```{r}
#| label: raw_data_export

#in this chunk a csv file containing the raw data will be generated, and uploaded for others to see and use. No filtering or cleaning has taken place so far, only joing the different datasets together to create the calluna hydraulics dataset.

write.csv(xylem_area_combined, "raw_data/DURIN_raw_LygraFire_field_trait_2025.csv")
```

# Plotting

```{r}
#| label: means and plot prepping

#adding means on different levels and adding statistics to visualize the data better

xylem_area_combined <- xylem_area_combined |>
  #creating means at the plant level
  group_by(plant) |>
  mutate(means_plant = mean(Area)) |>
  #creating means at the subset level
  group_by(source_file) |>
  mutate(means_subset = mean(Area)) |>
  group_by(treatment) |>
  mutate(means_treatment = mean(Area)) |>
  mutate(circ = (4*3.14)*(Area/(Perim.*Perim.))) |>
  ungroup()

xylem_area_combined$age_class <- factor(xylem_area_combined$age_class, levels = c("PIO", "BUI", "MAT"))

xylem_area_combined_circ <- xylem_area_combined |>
  filter(circ >= 0.7) |>
  group_by(treatment) |>
  mutate(means_treatment = mean(Area))

```

### Plant level xylem size vs subset damage

```{r}
#| label: xylem_size_damage_subset
#| message: false
#| warning: false

#simple visiualisation of plant level mean xylem size colored by extent of damage

ggplot(data = xylem_area_combined, mapping = aes(x = plant, y = Area, colour = as.character(as.numeric(subset_damage)))) + geom_point() + facet_wrap(vars(age_class), scales = "free_x") + scale_color_viridis_d()
```

### Plant level xylem mean size vs NiN

```{r}
#| label: xylem_mean_size_nin
#| message: false
#| warning: false

#simple visiualisation of plant level mean xylem size colored by nin type

ggplot(data = xylem_area_combined, mapping = aes(x = plant, y = means_plant, colour = nin_type)) + geom_jitter() + facet_wrap(vars(age_class), scales = "free") + scale_color_viridis_d()
```

### Xylem distribution changes unfiltered vs circularity \> 0.7

```{r}
#| label: xylem_mean_size_nin_subset
#| message: false
#| warning: false

#simple visiualisation of subset level mean xylem size colored by nin type

subset_nin_plot <- ggplot(data = xylem_area_combined, mapping = aes(x = treatment, y = Area, colour = nin_type)) + geom_jitter() + geom_violin() + facet_wrap(vars(age_class), scales = "free") + scale_color_viridis_d() + labs(title = "Unfiltered") + xlab(label = "NiN x Age class") + ylab("Xylem area in µm") + theme(legend.position = "none")

subset_nin_plot_filtered <- ggplot(data = xylem_area_combined_circ, mapping = aes(x = treatment, y = Area, colour = nin_type)) + geom_jitter() + geom_violin() + facet_wrap(vars(age_class), scales = "free") + scale_color_viridis_d() + labs(title = "Circularity > 0.7") + xlab(label = "NiN x Age class") + ylab("Xylem area in µm")

subset_nin_plot + subset_nin_plot_filtered
```

```{r}
#| label: xylem_density

xylem_density <- xylem_area_combined |>
  group_by(source_file) |>
  mutate(subset_xylem_total = sum(Area)) |>
  ungroup() |>
  mutate(xylem_density = subset_xylem_total/subset_area) |>
  group_by(plant) |>
  mutate(plant_mean_density = mean(xylem_density)) |>
  ungroup() |>
  group_by(treatment) |>
  mutate(treatment_mean_density = mean(xylem_density))


xylem_mean_plot <- ggplot(data = xylem_density) + geom_point(aes(x = age_class, y = means_treatment, color = nin_type)) + labs(x = "Age class", y = "Mean xylem vessel area") + guides(color = "none")

xylem_density_plot <-ggplot(data = xylem_density) + geom_point(aes(x = age_class, y = treatment_mean_density, color = nin_type)) + labs(x = "Age class", y = "Density (Total xylem area / subset area)") + guides(color = guide_legend("NiN type"))

xylem_mean_plot + xylem_density_plot

```

### Distribution of damage index 4 and 5, and the effect of circularity \> 0.7

```{r}
#creating new filtered datasets
xylem_area_damage_5 <- xylem_area_combined |>
  filter(subset_damage == 5)

xylem_area_damage_4 <- xylem_area_combined |>
  filter(subset_damage == 4)

xylem_area_damage_5_filtered <- xylem_area_combined_circ |>
  filter(subset_damage == 5)

xylem_area_damage_4_filtered <- xylem_area_combined_circ |>
  filter(subset_damage == 4)

#create a custom palette to use
palette_damage <- c("orange", "red")


plot_damage_5 <- ggplot() + 
  geom_jitter(data = xylem_area_combined,aes(x = treatment, y = Area), color = "grey") + geom_jitter(data = xylem_area_damage_4, aes(x = treatment, y = Area, color = as.character(as.numeric(subset_damage)))) + geom_jitter(data = xylem_area_damage_5, aes(x = treatment, y = Area, color = as.character(as.numeric(subset_damage)))) + geom_point(data = xylem_area_combined, aes(x = treatment, y = means_treatment), color = "blue") + labs(title = "Unfiltered") + ylab(label = "Xylem vessel area in µm") + xlab(label = "NiN x Age class") + theme(legend.position = "none") + scale_color_manual(values = palette_damage) + scale_y_continuous(breaks = seq(100,2000, by = 100))

plot_damage_5_filtered <- ggplot() + 
  geom_jitter(data = xylem_area_combined_circ,aes(x = treatment, y = Area), color = "grey") + geom_jitter(data = xylem_area_damage_4_filtered, aes(x = treatment, y = Area, color = as.character(as.numeric(subset_damage)))) + geom_jitter(data = xylem_area_damage_5_filtered, aes(x = treatment, y = Area, color = as.character(as.numeric(subset_damage)))) + geom_point(data = xylem_area_combined_circ, aes(x = treatment, y = means_treatment), color = "blue") + labs(title = "Circularity > 0.7") + ylab(label = "Xylem vessel area in µm") + xlab(label = "NiN x Age class") + scale_color_manual(values = palette_damage) + guides(color = guide_legend(title = "Subset damage index")) + scale_y_continuous(breaks = seq(100,2000, by = 100))

plot_damage_5 + plot_damage_5_filtered
```

### The effect of warped tissue

```{r}
#| label: warped vs unwarped

xylem_unwarped <- xylem_area_combined |>
  filter(warped == 0) |>
  group_by(treatment) |>
  mutate(means_treatment = mean(Area))

xylem_warped <- xylem_area_combined |>
  filter(warped == 1) |>
  group_by(treatment) |>
  mutate(means_treatment = mean(Area))

plot_unwarped <- ggplot(data = xylem_unwarped) + geom_jitter(aes(x = treatment, y = Area), color = "grey") + geom_point(aes(x = treatment, y = means_treatment), color = "blue") + labs(title = "Unwarped")

plot_warped <- ggplot(data = xylem_warped) + geom_jitter(aes(x = treatment, y = Area), color = "grey") + geom_point(aes(x = treatment, y = means_treatment), color = "blue") + labs(title = "Warped")

plot_unwarped + plot_warped
```

Warps have a lower average xylem size by around a fixed value of 100 square µm. Note that the xylem vessels themselves are not necessarily warped, but the tissue they were found in is. Regardless, the absolute size of xylem vessels seem to decrease in warped tissue.

### The effect of folds

```{r}
#| label: the effect of folds

xylem_folds0 <- xylem_area_combined |>
  filter(folds == 0) |>
  group_by(treatment) |>
  mutate(means_folds = mean(Area))

xylem_folds1 <- xylem_area_combined |>
  filter(folds == 1) |>
  group_by(treatment) |>
  mutate(means_folds = mean(Area))

xylem_folds2 <- xylem_area_combined |>
  filter(folds == 2) |>
  group_by(treatment) |>
  mutate(means_folds = mean(Area))

xylem_folds3 <- xylem_area_combined |>
  filter(folds == 3) |>
  group_by(treatment) |>
  mutate(means_folds = mean(Area))

plot_folds_control <- ggplot(data = xylem_area_combined) + geom_jitter(aes(x = treatment, y = Area), color = "grey") + geom_point(aes(x = treatment, y = means_treatment), color = "blue") + labs(title = "Control")+ scale_y_continuous(breaks = seq(100,2000, by = 100))

plot_folds0 <- ggplot(data = xylem_folds0) + geom_jitter(aes(x = treatment, y = Area), color = "grey") + geom_point(aes(x = treatment, y = means_folds), color = "blue") + labs(title = "Folds = 0")  + scale_y_continuous(breaks = seq(100,2000, by = 100))

plot_folds1 <- ggplot(data = xylem_folds1) + geom_jitter(aes(x = treatment, y = Area), color = "grey") + geom_point(aes(x = treatment, y = means_folds), color = "blue") + labs(title = "Folds = 1")  + scale_y_continuous(breaks = seq(100,2000, by = 100))

plot_folds2 <- ggplot(data = xylem_folds2) + geom_jitter(aes(x = treatment, y = Area), color = "grey") + geom_point(aes(x = treatment, y = means_folds), color = "blue") + labs(title = "Folds = 2")  + scale_y_continuous(breaks = seq(100,2000, by = 100))

plot_folds3 <- ggplot(data = xylem_folds3) + geom_jitter(aes(x = treatment, y = Area), color = "grey") + geom_point(aes(x = treatment, y = means_folds), color = "blue") + labs(title = "Folds = 3")  + scale_y_continuous(breaks = seq(100,2000, by = 100))


plot_folds_control + plot_folds0 + plot_folds1 + plot_folds2 + plot_folds3
```

Folds \>= 2 seem to decrease the average size of xylem, while folds = 1 is similar to folds = 0, with the exception of MAT-BH, which is a bit higher.

```{r}
#| label: the effect of folds, but cumulative


xylem_folds1_cumulative <- xylem_area_combined |>
  filter(folds < 3) |>
  group_by(treatment) |>
  mutate(means_folds = mean(Area))

xylem_folds2_cumulative <- xylem_area_combined |>
  filter(folds < 2) |>
  group_by(treatment) |>
  mutate(means_folds = mean(Area))

xylem_folds3_cumulative <- xylem_area_combined |>
  filter(folds < 1) |>
  group_by(treatment) |>
  mutate(means_folds = mean(Area))

plot_folds_control_cumulative <- ggplot(data = xylem_area_combined) + geom_jitter(aes(x = treatment, y = Area), color = "grey") + geom_point(aes(x = treatment, y = means_treatment), color = "blue") + labs(title = "Control")+ scale_y_continuous(breaks = seq(100,2000, by = 100))

plot_folds1_cumulative <- ggplot(data = xylem_folds1_cumulative) + geom_jitter(aes(x = treatment, y = Area), color = "grey") + geom_point(aes(x = treatment, y = means_folds), color = "blue") + labs(title = "Folds < 3")  + scale_y_continuous(breaks = seq(100,2000, by = 100))

plot_folds2_cumulative <- ggplot(data = xylem_folds2_cumulative) + geom_jitter(aes(x = treatment, y = Area), color = "grey") + geom_point(aes(x = treatment, y = means_folds), color = "blue") + labs(title = "Folds < 2")  + scale_y_continuous(breaks = seq(100,2000, by = 100))

plot_folds3_cumulative <- ggplot(data = xylem_folds3_cumulative) + geom_jitter(aes(x = treatment, y = Area), color = "grey") + geom_point(aes(x = treatment, y = means_folds), color = "blue") + labs(title = "Folds < 1")  + scale_y_continuous(breaks = seq(100,2000, by = 100))


plot_folds_control_cumulative + plot_folds1_cumulative + plot_folds2_cumulative + plot_folds3_cumulative
```

Removing folds does not seem to impact average xylem size visually. At "Folds \< 1", there is almost 3000 observations removed.

### Combining circularity, warps, and folds

```{r}
#| label: circuarity, warps, and folds

xylem_combined_factors <- xylem_area_combined |>
  filter(circ >= 0.70) |>
  filter(warped == 0) |>
  filter(folds == 0) |>
  group_by(treatment) |>
  mutate(means_treatment = mean(Area))


plot_control <- ggplot(data = xylem_area_combined) + geom_jitter(aes(x = treatment, y = Area), color = "grey") + geom_point(aes(x = treatment, y = means_treatment), color = "blue") + labs(title = "Control")  + scale_y_continuous(breaks = seq(100,2000, by = 100))

plot_combined_factors <- ggplot(data = xylem_combined_factors) + geom_jitter(aes(x = treatment, y = Area), color = "grey") + geom_point(aes(x = treatment, y = means_treatment), color = "blue") + labs(title = "Combined factors")  + scale_y_continuous(breaks = seq(100,2000, by = 100))

plot_control + plot_combined_factors
```

BUI-FH, BUI-TH, PIO-BH, and PIO-TH all increased after filtering by circularity \> 0.70, warps = 0, and folds = 0. The rest of the averages stayed basically the same.\
\

```{r}
#| label: subset average damage

#this chunk showcases the average xylem size for different subset damages

avg_damage <- xylem_area_combined |>
  group_by(subset_damage, treatment) |>
  mutate(mean_damage = mean(Area))


ggplot(data = avg_damage) + geom_jitter(aes(x = treatment, y = Area), color = "grey") + geom_point(aes(x = treatment, y = mean_damage, color = subset_damage))
```

Higher damage seem to decrease the average xylem size, so by filtering away damaged subsets it is likely that the average xylem size would increase

```{r}
circ_vs_area <- xylem_area_combined |>
  mutate(circ = (4*3.14)*(Area/(Perim.*Perim.)))

circ_vs_area <- circ_vs_area |>
  rename_at("Perim.", ~ "Perim")

circ_vs_area_plot <- ggplot(data = circ_vs_area) + geom_point(aes(x = circ, y = Area, color = Feret), alpha = 0.4) + labs(x = "Circularity")

feret_vs_area_plot <- ggplot(data = circ_vs_area) + geom_point(aes(x = Feret, y = Area, color = circ), alpha = 0.4) + labs(x = "Feret diameter") + guides(color = guide_legend("Circularity"))

perim_vs_area_plot <- ggplot(data = circ_vs_area) + geom_point(aes(x = Perim, y = Area, color = circ), alpha = 0.4) + labs(x = "Circumference") + guides(color = guide_legend("Circularity"))

circ_vs_area_plot + feret_vs_area_plot + perim_vs_area_plot
```

## Mixed effect models

```{r}

model_mixed <- lmer(log(Area) ~ nin_type * age_class + (1 | plant), data = xylem_area_combined)

model_mixed_combined <- lmer(log(Area) ~ nin_type * age_class + (1 | plant), data = xylem_combined_factors)

plot(model_mixed)
plot(model_mixed_combined)

summary(model_mixed)
summary(model_mixed_combined)
```
